---
title: "Bill.com Forcast"
author: "Lynn Niu, James Li, Bosco Zhu "
date: "2021/2/6"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Cleaning
```{r}
library(magrittr)
library(dplyr)
library(lubridate)
library(ggplot2)
library(readr)
df <- read_csv("C:/R Dataset/Purchase_Card_Transactions.csv")
names(df)[1] = "OBJECTID"
# Remove NAs
df[df == "N/A"] = NA
df[df == ""] = NA
```

# Exploratory Analysis
```{r}
# Density of transaction amount
ggplot(df, aes(x=TRANSACTION_AMOUNT)) + geom_density() + xlim(-5000,5000)

# Transaction distribution by industry
t_list <- as.data.frame(rev(sort(table(df$MCC_DESCRIPTION))))
plot(t_list$Freq, type = "line", ylab = "Aggregate Transaction Amount", xlab = "Aggregate Transaction Types")
# Industries with 5000+ Transaction Volumes
industry <- subset(t_list, Freq>5000)
# Composition
comp <- data.frame(industry = c("Consumer Staples","Business Service", "Industrials", "Technology", "Donation", "Education", "Government"), transaction_num = c(
  sum(industry[c(1,2,3,5,11,16),2]),
  sum(industry[c(4,9,14,15,17),2]),
  sum(industry[c(6,13),2]),
  sum(industry[c(7,18),2]),
  sum(industry[8,2]),
  sum(industry[10,2]),
  sum(industry[12,2]))
)
pie(comp$transaction_num, labels = comp$industry, main="Transaction Distribution among Industries ")
```




# Spending Forcast for the agencies
* remove large outliers

# Data Reorganization
```{r}
# Reformatting
df %>% 
  mutate(TRANSACTION_DATE = format(as_date(ymd_hms(TRANSACTION_DATE)), "%Y-%m"))%>%
  group_by(AGENCY, TRANSACTION_DATE) %>%
  summarise(TRANSACTION_AMOUNT_SUM = sum(TRANSACTION_AMOUNT),
            MONTHLY_COUNTS = n()) -> df_agency_0
df_agency_0 %>%
    mutate(month = substring(TRANSACTION_DATE,6,7), 
         year = substring(TRANSACTION_DATE,1,4)) -> df_agency
df_agency

# Remove Outliers
plot(df_agency$TRANSACTION_AMOUNT_SUM)
df_agency[which(df_agency$TRANSACTION_AMOUNT_SUM>200000),]
df_agency = df_agency[-which(df_agency$TRANSACTION_AMOUNT_SUM>200000),]
plot(df_agency$TRANSACTION_AMOUNT_SUM)
df_agency$TRANSACTION_AMOUNT_SUM
```


# MLR Inference
```{r}
# MLR Inference: not a good fit
fit.lm = lm(TRANSACTION_AMOUNT_SUM ~ MONTHLY_COUNTS + month + year, df_agency)
summary(fit.lm)
par(mfrow = c(2,2))
plot(fit.lm)
```

# Training and Testing Datasets
```{r}
size <- nrow(df_agency)
sampler <- sample(as.vector(c(rep(TRUE,0.7*size+1),rep(FALSE,0.3*size))))
training <- df_agency[sampler,]
testing <- df_agency[!sampler,]
```

# GAM Model
```{r}
library(mgcv)
fit.gam = gam(TRANSACTION_AMOUNT_SUM ~ MONTHLY_COUNTS + month + year, data = training)
summary(fit.gam)
yhat.gam = predict(fit.gam, newdata = testing)
(MSE.gam = mean((yhat.gam - testing$TRANSACTION_AMOUNT_SUM)^2))
plot(testing$TRANSACTION_AMOUNT_SUM, main = "General Additive Model: Actual(Red) vs Predicted(Green)", type = "l", col = "green")
lines(yhat.gam, col="brown")
```
## Decision Tree Model
```{r, fig.align='center'}
training
library(ISLR)
library(tree)
set.seed(100)
fit.tree = tree(TRANSACTION_AMOUNT_SUM~., training[3:6])
plot(fit.tree)
text(fit.tree, pretty = 0)
#summary(fit.tree)
yhat.tree = predict(fit.tree, newdata =testing)
(MSE.tree = mean((yhat.tree - testing$TRANSACTION_AMOUNT_SUM)^2))
plot(testing$TRANSACTION_AMOUNT_SUM, main = "Decision Tree Model: Actual(Red) vs Predicted(Green)", type = "line", col = "green") 
lines(yhat.gam, col="brown")
```

# Bagging
```{r, fig.align='center'}
library(randomForest)
#bagging
fit.bag = randomForest(TRANSACTION_AMOUNT_SUM~., training[3:6], mtry=3, 
                      importance =TRUE, ntree = 500)
yhat.bag = predict(fit.bag, newdata = testing)
(MSE.bag = mean((yhat.bag - testing$TRANSACTION_AMOUNT_SUM)^2))
```
# Ensemble Method: Boosting
```{r, fig.align='center'}
library(gbm)


training$month <- as.numeric(training$month)
training$year <- as.numeric(training$year)
training$MONTHLY_COUNTS <- as.numeric(training$MONTHLY_COUNTS)


set.seed(100)
fit.boost = gbm(TRANSACTION_AMOUNT_SUM~., training[3:6], 
                distribution = "gaussian", n.trees =500, 
                interaction.depth = 5)
summary(fit.boost)
yhat.boost=predict(fit.boost, newdata =testing, n.trees =500)
(MSE.boosting = mean((yhat.boost -testing$TRANSACTION_AMOUNT_SUM)^2))
plot(testing$TRANSACTION_AMOUNT_SUM, main = "Boosting Regression: Actual(Red) vs Predicted(Green)", type = "line", col = "green")
lines(yhat.gam, col="brown")
```





# Receivable forecasting for the vendor
* remove large outliers
* one_time: transaction shows up once
* some_times: transaction shows up 2 to 8 times
* many_times: transaction shows up more than 8 times

# Data Reorganization
```{r}
# regroup VENDOR_NAME into less categories
# using grouped monthly data
# group by month
df %>% 
  mutate(TRANSACTION_DATE = format(as_date(ymd_hms(TRANSACTION_DATE)), "%Y-%m"))%>%
  group_by( VENDOR_NAME, TRANSACTION_DATE) %>%
  summarise(TRANSACTION_AMOUNT_SUM = sum(TRANSACTION_AMOUNT, na.omit = TRUE),
            MONTHLY_COUNTS = n()) %>% 
  mutate(month = as.numeric(substring(TRANSACTION_DATE,6,7)), 
         year = substring(TRANSACTION_DATE,1,4)) %>% 
  add_count(VENDOR_NAME)  -> df_vendor


#check outliers: monthly transaction exceeds $1,000,000
par(mfrow = c(2,2))
plot(df_vendor$TRANSACTION_AMOUNT_SUM)
df_vendor[which(df_vendor$TRANSACTION_AMOUNT_SUM>200000),]
df_vendor = df_vendor[-which(df_vendor$TRANSACTION_AMOUNT_SUM>200000),]
plot(df_vendor$TRANSACTION_AMOUNT_SUM)
dev.off()

#Exploration:
df_vendor %>% 
  group_by(VENDOR_NAME) %>% 
  summarise(n = n()) -> df_occurrence

df_occurrence

sum(df_occurrence$n == 1) #number of vendor transaction occurrence=1
sum(df_occurrence$n>1 & df_occurrence$n<=10) #between 1 and 8
sum(df_occurrence$n > 10) #greater than 8


df_vendor$n = ifelse(df_vendor$n == 1, 
                     "one_time", ifelse(df_vendor$n > 10, "many_times", "some_times"))
names(df_vendor)[7] = "occurrence"
df_vendor$occurrence = as.factor(df_vendor$occurrence)
df_vendor$year = as.factor(df_vendor$year)
```

# MLR Inference
```{r}
fit.lm = lm(TRANSACTION_AMOUNT_SUM ~ MONTHLY_COUNTS + month + year + occurrence, df_vendor)
summary(fit.lm)
par(mfrow = c(2,2))
plot(fit.lm)
dev.off()
```
*occurrence = one_time, some_time, MONTHLY_COUNTS, month are significant predictors
*R-sqr is small, poor fit

# Set training and testing data sets
Each time we use training set to train the model and use testing set to assess model and compute MSE to compare models.
```{r}
set.seed(100)
nrow(df_vendor)
test = sample(nrow(df_vendor),10000)
train = -test
df.test = df_vendor[test,]
df.train = df_vendor[train,]
```

# General Additive Model
```{r}
library(mgcv)
fit.gam = gam(TRANSACTION_AMOUNT_SUM ~ s(MONTHLY_COUNTS) + month + year + occurrence, data = df.train)
summary(fit.gam)
yhat.gam = predict(fit.gam, newdata = df.test)
(MSE.gam = mean((yhat.gam - df.test$TRANSACTION_AMOUNT_SUM)^2))
plot(df.test$TRANSACTION_AMOUNT_SUM, main = "General Additive Model: Actual(Red) vs Predicted(Green)", type ="l", col = "green")
lines(yhat.gam, col="brown")
```
Large MSE, bad fit.



# Decision Tree Model
```{r, fig.align='center'}
library(ISLR)
library(tree)
set.seed(100)
fit.tree = tree(TRANSACTION_AMOUNT_SUM~., df.train[3:7])
plot(fit.tree)
text(fit.tree, pretty = 0)
#summary(fit.tree)
yhat.tree = predict(fit.tree, newdata =df.test)
(MSE.tree = mean((yhat.tree - df.test$TRANSACTION_AMOUNT_SUM)^2))
plot(df.test$TRANSACTION_AMOUNT_SUM, main = "Decision Tree Model: Actual(Red) vs Predicted(Green)", type = "l", col = "green")
lines(yhat.gam, col="brown")
```

# Bagging
```{r, fig.align='center'}
library(randomForest)
#bagging
fit.bag = randomForest(TRANSACTION_AMOUNT_SUM~., df.train[3:7], mtry=4, 
                      importance =TRUE, ntree = 5)
yhat.bag = predict(fit.bag, newdata = df.test)
(MSE.bag = mean((yhat.bag - df.test$TRANSACTION_AMOUNT_SUM)^2))

```

# Ensemble Method: Boosting
```{r, fig.align='center'}
library(gbm)

set.seed(100)

fit.boost = gbm(TRANSACTION_AMOUNT_SUM~., df.train[3:7], 
                distribution = "gaussian", n.trees =500, 
                interaction.depth = 5)
summary(fit.boost)
yhat.boost=predict(fit.boost, newdata =df.test, n.trees =500)
(MSE.boosting = mean((yhat.boost -df.test$TRANSACTION_AMOUNT_SUM)^2))
plot(df.test$TRANSACTION_AMOUNT_SUM, main = "Boosting Model: Actual(Red) vs Predicted(Green)", col = "green")
lines(yhat.gam, col="brown")

```


