---
title: "bill.com forecasting"
author: "Lynn Niu"
date: "2021/2/6"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Cleaning
```{r}
library(magrittr)
library(dplyr)
library(lubridate)
library(ggplot2)

df = read.csv("Purchase_Card_Transactions.csv")
names(df)[1] = "OBJECTID"
#unique(df$AGENCY)
#unique(df$VENDOR_NAME)
#unique(df$MCC_DESCRIPTION)
#unique(df$VENDOR_STATE_PROVINCE)


df[df == "N/A"] = NA
df[df == ""] = NA

# density of transaction amount
ggplot(df, aes(x=TRANSACTION_AMOUNT)) + geom_density() + xlim(-5000,5000)

# df %>% 
#   mutate(TRANSACTION_DATE = format(as_date(ymd_hms(TRANSACTION_DATE)), "%Y-%m"))%>%
#   group_by( AGENCY, TRANSACTION_DATE) %>%
#   summarise(TRANSACTION_AMOUNT_SUM = sum(TRANSACTION_AMOUNT),
#             TRANSACTION_COUNT = n()) -> df_agency

#Group by month
df %>% 
  mutate(TRANSACTION_DATE = format(as_date(ymd_hms(TRANSACTION_DATE)), "%Y-%m"))%>%
  group_by( VENDOR_NAME, TRANSACTION_DATE) %>%
  summarise(TRANSACTION_AMOUNT_SUM = sum(TRANSACTION_AMOUNT, na.omit = TRUE),
            MONTHLY_COUNTS = n()) %>% 
  mutate(month = as.numeric(substring(TRANSACTION_DATE,6,7)), 
         year = substring(TRANSACTION_DATE,1,4)) %>% 
  add_count(VENDOR_NAME)  -> df_vendor


```

# Spend forecasting for the agencies
```{r}

```

# Receivable forecasting for the vendor
* remove large outliers
* one_time: transaction shows up once
* some_times: transaction shows up 2 to 8 times
* many_times: transaction shows up more than 8 times

```{r}
# regroup VENDOR_NAME into less categories
# using grouped monthly data

#check outliers: monthly transaction exceeds $1,000,000
par(mfrow = c(1,2))
plot(df_vendor$TRANSACTION_AMOUNT_SUM)
df_vendor[which(df_vendor$TRANSACTION_AMOUNT_SUM>100000),]
df_vendor = df_vendor[-which(df_vendor$TRANSACTION_AMOUNT_SUM>100000),]
plot(df_vendor$TRANSACTION_AMOUNT_SUM)
dev.off()

#Exploration:
df_vendor %>% 
  group_by(VENDOR_NAME) %>% 
  summarise(n = n()) -> df_occurrence

sum(df_occurrence$n == 1) #number of vendor transaction occurrence=1
sum(df_occurrence$n>1 & df_occurrence$n<=10) #between 1 and 8
sum(df_occurrence$n > 10) #greater than 8


df_vendor$n = ifelse(df_vendor$n == 1, 
                     "one_time", ifelse(df_vendor$n > 10, "many_times", "some_times"))
names(df_vendor)[7] = "occurrence"
df_vendor$occurrence = as.factor(df_vendor$occurrence)
df_vendor$year = as.factor(df_vendor$year)

```

## MLR
```{r}
fit.lm = lm(TRANSACTION_AMOUNT_SUM ~ MONTHLY_COUNTS + month + year + occurrence, df_vendor)
summary(fit.lm)
par(mfrow = c(2,2))
plot(fit.lm)
dev.off()
plot(df_vendor$TRANSACTION_AMOUNT_SUM, main = "MLR fitted (red)")
points(predict(fit.lm), col="brown")
```
*occurrence = one_time, some_time, MONTHLY_COUNTS, month are significant predictors
*R-sqr is small, poor fit

## Set training, testing sets (10000)
Each time we use training set to train the model and use testing set to assess model and compute MSE to compare models.
```{r}
set.seed(100)
test = sample(nrow(df_vendor),10000)
train = -test
df.test = df_vendor[test,]
df.train = df_vendor[train,]
```

## GAM
```{r}
library(mgcv)
fit.gam = gam(TRANSACTION_AMOUNT_SUM ~ s(MONTHLY_COUNTS) + month + year + occurrence, data = df.train)
summary(fit.gam)
yhat.gam = predict(fit.gam, newdata = df.test)
(MSE.gam = mean((yhat.gam - df.test$TRANSACTION_AMOUNT_SUM)^2))
plot(df.test$TRANSACTION_AMOUNT_SUM, main = "GAM fitted (red)")
points(yhat.gam, col="brown")
```
Large MSE, bad fit.

## Decision Tree
```{r, fig.align='center'}
library(ISLR)
library(tree)
set.seed(100)
fit.tree = tree(TRANSACTION_AMOUNT_SUM~., df.train[3:7])
plot(fit.tree)
text(fit.tree, pretty = 0)
#summary(fit.tree)
yhat.tree = predict(fit.tree, newdata =df.test)
(MSE.tree = mean((yhat.tree - df.test$TRANSACTION_AMOUNT_SUM)^2))
plot(df.test$TRANSACTION_AMOUNT_SUM, main = "Decision Tree fitted (red)")
points(yhat.gam, col="brown")
```

## Bagging
```{r, fig.align='center'}
library(randomForest)
#bagging
fit.bag = randomForest(TRANSACTION_AMOUNT_SUM~., df.train[3:7], mtry=4, 
                      importance =TRUE, ntree = 5)
yhat.bag = predict(fit.bag, newdata = df.test)
(MSE.bag = mean((yhat.bag - df.test$TRANSACTION_AMOUNT_SUM)^2))

```

#### Boosting
```{r, fig.align='center'}
library(gbm)

set.seed(100)

fit.boost = gbm(TRANSACTION_AMOUNT_SUM~., df.train[3:7], 
                distribution = "gaussian", n.trees =500, 
                interaction.depth = 5)
summary(fit.boost)
yhat.boost=predict(fit.boost, newdata =df.test, n.trees =500)
(MSE.boosting = mean((yhat.boost -df.test$TRANSACTION_AMOUNT_SUM)^2))
plot(df.test$TRANSACTION_AMOUNT_SUM, main = "Boosting fitted (red)")
points(yhat.gam, col="brown")

```
